---
title: "FIFA World Cup Attendance & Climate Analysis"
author: "Nihar Khadawala"
output: html_document
date: "2026-05-10"
---

## Introduction

The FIFA World Cup is one of the world's most widely watched sporting events, held every four years in a different host country. This project investigates whether climate and tournament timing are associated with stadium attendance.

The analysis focuses on two questions:

1. Is weather associated with average attendance per game?
2. Does the month in which a tournament begins relate to average attendance?

Two datasets are used. The first contains total and average attendance figures for FIFA World Cups from 1930 to 2018. The second contains tournament dates and general weather descriptions. Because the number of matches varied substantially across tournaments, average attendance per game is used as the primary measure throughout.

## Data Sources

- **Statista:** FIFA World Cup attendance figures, 1930–2018.
- **Reuters / Yahoo Sports:** Tournament dates and general weather descriptions.

The Statista data was manually entered into a tibble, since the underlying dataset was viewable but not directly downloadable. The Reuters/Yahoo Sports data was collected through web scraping and then cleaned and structured for analysis.

**Statista:** https://www.statista.com/statistics/264441/number-of-spectators-at-football-world-cups-since-1930/

**Reuters / Yahoo Sports:** https://sports.yahoo.com/soccer-dates-weather-previous-world-cups-130013936--sow.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(rvest)
library(ggbeeswarm)
```

## Data Source 1: Statista Attendance Figures

The Statista dataset contains total and average per-game attendance for all 21 World Cups held between 1930 and 2018. The source page was viewable but paywalled for direct download, so each data point was manually entered into the tibble below.

```{r statista-data}
attendance <- tibble(
  year = c(2018, 2014, 2010, 2006, 2002, 1998, 1994, 1990, 1986,
            1982, 1978, 1974, 1970, 1966, 1962, 1958, 1954, 1950,
            1938, 1934, 1930),
  host = c("Russia", "Brazil", "South Africa", "Germany",
            "Japan / South Korea", "France", "USA", "Italy", "Mexico",
            "Spain", "Argentina", "Germany", "Mexico", "England",
            "Chile", "Sweden", "Switzerland", "Brazil", "France",
            "Italy", "Uruguay"),
  total_attendance = c(3031768, 3441450, 3167984, 3367000, 2724604,
                       2859234, 3568567, 2527348, 2407431, 1856277,
                       1610215, 1774022, 1673975, 1614677, 776000,
                       868000, 943000, 1337000, 483000, 395000, 434000),
  avg_attendance   = c(47371, 53772, 49499, 52609, 42571, 44676,
                       68626, 48411, 46297, 35698, 42374, 46685,
                       52312, 50459, 24250, 24800, 36269, 60773,
                       26833, 23235, 24139)
)
```

The number of games played has changed significantly over the tournament's history — early World Cups had far fewer teams and matches, so total attendance figures are not directly comparable across years. For that reason, **average attendance per game** is the more reliable metric and is used for comparison throughout this report.

### Attendance Over Time

```{r attendance-table}
attendance |>
  arrange(year) |>
  rename(
    Year                      = year,
    Host                      = host,
    `Total Attendance`        = total_attendance,
    `Avg Attendance per Game` = avg_attendance
  ) |>
  print(n = Inf)
```

```{r attendance-plot}
attendance |>
  ggplot(aes(year, avg_attendance)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(color = "blue", size = 2.5) +
  scale_x_continuous(breaks = attendance$year) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Average Attendance per Game at FIFA World Cups (1930–2018)",
    x     = "Year",
    y     = "Average Attendance per Game"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Attendance was lower in the earliest tournaments, then peaked sharply in 1950 (Brazil) and again in 1994 (United States). Since then, attendance has remained relatively stable in the 45,000–55,000 range. Note that the 1942 and 1946 World Cups are absent from the data — both were cancelled due to World War II.

The lowest average attendance occurred in 1962 (Chile), and the highest in 1994 (United States). The 1994 peak lines up with the large-capacity American football stadiums used that year, while the 1962 low corresponds to a 9.5-magnitude earthquake that struck Chile in 1960 and damaged infrastructure ahead of the tournament.

Sources:
https://www.foxsports.com/stories/soccer/1994-world-cup-united-states-history-host
https://en.wikipedia.org/wiki/1962_FIFA_World_Cup

## Data Source 2: Weather and Tournament Dates (Reuters / Yahoo Sports)

The Reuters article provides start and end dates and a brief weather description for each World Cup from 1930 to 2010. Unlike the Statista data, this required web scraping and cleaning.

```{r scraping-permission, echo=TRUE}
url <- "https://sports.yahoo.com/soccer-dates-weather-previous-world-cups-130013936--sow.html"
robotstxt::paths_allowed(url)
```

With scraping permission confirmed, the page was read with `read_html()`, and all paragraph text was extracted with `html_elements("p")` and `html_text2()`.

```{r scrape-reuters, cache=TRUE}
page <- read_html(url)

raw_paragraphs <- page |>
  html_elements("p") |>
  html_text2()
```

The article is written as one continuous paragraph with no line breaks between tournament entries. Each entry follows the format:

```
1930 Uruguay July 13-July 30 Warm wintry weather
1934 Italy May 27-June 10 Warm and sunny most of time
```

### Cleaning the Weather Data

To parse this, each tournament entry was isolated by inserting a `|` separator before every four-digit year, splitting the text on that separator, and extracting the year, start month, and weather description.

```{r isolate-lines}
article_text <- raw_paragraphs[[2]]

data_section <- article_text |>
  str_extract("Dates and overall weather at previous finals: .+")

weather_lines <- data_section |>
  str_replace_all("(\\d{4} )", "|\\1") |>
  str_split("\\|") |>
  _[[1]] |>
  str_subset("^\\d{4}")
```

```{r clean-weather-fields}
weather <- tibble(line = weather_lines) |>
  mutate(
    year = str_extract(line, "^\\d{4}") |> as.integer(),

    start_month = str_extract(line, "[A-Z][a-z]+ \\d") |>
                    str_remove(" \\d"),

    weather = str_replace(
                line,
                "^\\d{4}\\s+[\\w/. ]+?\\s+[A-Z][a-z]+ \\d{1,2}-[A-Z][a-z]+ \\d{1,2}\\s+",
                ""
              ) |>
              str_trim()
  ) |>
  select(year, start_month, weather)
```

```{r month-to-numeric}
weather <- weather |>
  mutate(month_num = match(start_month, month.name))
```

The cleaned weather table contains 19 rows, one for each World Cup covered by the article. The 2014 and 2018 tournaments are not mentioned in the article and are therefore absent.

```{r weather-table}
weather |>
  arrange(year) |>
  rename(
    Year                  = year,
    `Start Month`         = start_month,
    `Month Number`        = month_num,
    `Weather Description` = weather
  ) |>
  print(n = Inf)
```

## Merging the Datasets

Since both tibbles share a `year` column, they were merged using `left_join()`, keeping all 21 rows from the attendance data. The 2014 and 2018 entries have `NA` for the weather-related columns.

```{r join-data}
world_cup <- attendance |>
  left_join(weather, by = "year")
```

```{r save-csv, eval=FALSE}
world_cup |>
  write_csv("world_cup_tidy.csv")
```

## Analysis

### Weather Categories and Attendance

The weather descriptions in the article are not pre-categorized, but they follow recognizable patterns. Using `str_detect()` and `case_when()`, each tournament was assigned to one of three categories: *Cold / Wet*, *Variable*, or *Hot / Sunny*.

```{r weather-category}
world_cup <- world_cup |>
  mutate(
    weather_category = case_when(
      str_detect(weather, regex("cold|rain|wintry|wet", ignore_case = TRUE))
        ~ "Cold / Wet",
      str_detect(weather, regex("changeable|variable|humid", ignore_case = TRUE))
        ~ "Variable",
      str_detect(weather, regex("hot|sunny|warm|pleasant", ignore_case = TRUE))
        ~ "Hot / Sunny",
      !is.na(weather) ~ "Other",
      TRUE            ~ NA_character_
    )
  )
```

```{r weather-boxplot}
world_cup |>
  filter(!is.na(weather_category)) |>
  ggplot(aes(weather_category, avg_attendance,
             group = weather_category, fill = weather_category)) +
  geom_boxplot() +
  geom_beeswarm() +
  scale_y_continuous(labels = scales::label_comma()) +
  ylab("Average Attendance per Game") +
  xlab("Weather Category") +
  ggtitle("Average Attendance per Game by Weather Category") +
  theme_bw()
```

The Hot / Sunny category has the highest median attendance (~46,000), followed by Cold / Wet (~44,000) and Variable (~39,000). Warmer weather is associated with a modest increase in attendance relative to the other categories, but all three groups show considerable overlap, making it difficult to conclude that weather alone drives attendance.

```{r weather-summary-table}
world_cup |>
  filter(!is.na(weather_category)) |>
  group_by(weather_category) |>
  summarize(
    n              = n(),
    mean_avg_att   = round(mean(avg_attendance)),
    median_avg_att = round(median(avg_attendance)),
    .groups = "drop"
  ) |>
  rename(
    `Weather Category`      = weather_category,
    `N Tournaments`         = n,
    `Mean Avg Attendance`   = mean_avg_att,
    `Median Avg Attendance` = median_avg_att
  )
```

This summary table confirms the pattern from the boxplot: mean attendance is highest for warmer weather, followed by cooler and then variable conditions. With a sample of only 19 tournaments, though, this is far from conclusive — a larger sample would support more reliable inference.

### Tournament Timing and Attendance

All World Cups in this dataset started in May, June, or July. Even within that narrow window, it's worth checking whether the specific start month relates to attendance. The plot below shows average per-game attendance against start month for the 19 tournaments covered by the Reuters data.

```{r timing-plot, message=FALSE}
world_cup |>
  filter(!is.na(month_num)) |>
  ggplot(aes(month_num, avg_attendance)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  scale_x_continuous(breaks = 5:7,
                     labels = c("May", "June", "July")) +
  scale_y_continuous(labels = scales::label_comma()) +
  ylab("Average Attendance per Game") +
  xlab("Start Month") +
  ggtitle("Tournament Start Month vs. Average Attendance per Game") +
  theme_bw()
```

The trend line shows a slight positive slope — tournaments starting later in the summer tend to have marginally higher average attendance — but the line is close to flat, suggesting any relationship is likely noise rather than a true effect. A linear regression confirms this.

```{r regression}
lm_fit <- world_cup |>
  filter(!is.na(month_num)) |>
  lm(avg_attendance ~ month_num, data = _)

summary(lm_fit)
```

The p-value of 0.7234 is well above 0.05, indicating the relationship is not statistically significant, and an R-squared of 0.00756 shows start month explains less than 1% of the variation in attendance.

## Conclusion

This investigation into climate and tournament timing yielded a few key insights. Weather shows a slight association with attendance — hot and sunny conditions draw modestly larger crowds than cold, wet, or variable weather — though the effect is small relative to the overlap between categories. Tournament start month shows no meaningful relationship with attendance, as confirmed by the regression. Overall, spectator turnout appears to depend on more than weather or timing alone: stadium capacity (e.g., large NFL stadiums in the 1994 US tournament) and external shocks (e.g., the Chile earthquake ahead of the 1962 tournament) both appear to have played a larger role.

Future work could incorporate additional variables such as host country GDP, population, or ticket prices, enabling a multiple regression approach that controls for confounding factors and better isolates the effect of any single variable on attendance.
